---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "gauthier, danielle"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
# Summary statistics for all variables
summary(mydata)

# Frequency table of CLASS by RINGS
freq_table <- table(mydata$CLASS, mydata$RINGS)

# View the table
freq_table
length(freq_table)
kable(freq_table)


```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: (Enter your answer here.)***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
# Create frequency table of SEX by CLASS
sex_class_table <- table(mydata$SEX, mydata$CLASS)

# Add marginal totals
sex_class_margins <- addmargins(sex_class_table)

# View the table with margins
sex_class_margins

# Barplot of SEX by CLASS (excluding margins)
barplot(sex_class_table,
        beside = TRUE,
        col = c("skyblue", "lightgreen", "salmon"),
        legend = rownames(sex_class_table),
        main = "Counts by SEX and CLASS",
        xlab = "CLASS",
        ylab = "Count")


```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: 
There are five classes: A1 through A5, each associated with distinct ranges of RINGS (which serve for age). The distribution is segmented by age group, for example, A1 abalones all have low RINGS (3–6), while A5 abalones are in the highest RINGS range (13–25). This pattern shows no overlap between classes in terms of RINGS values. Each CLASS is essentially an age category rather than a random label.

The table shows complete separation by age: no A1 beyond 6 rings, no A2 below 7 or above 8 rings, no A3 below 9 or above 10 rings, and so on. Class sizes vary dramatically, mid-age classes (A2 and A3) have much larger counts than the youngest (A1) or oldest (A5). The older age group (A5) is rare, especially beyond 20 rings, suggesting either high mortality or reduced sampling frequency in older abalones.***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)  # for reproducibility
sample_indices <- sample(1:1036, size = 200)
work <- mydata[sample_indices, ]

plot(work[, 2:6],
     main = "Scatterplot Matrix of Selected Continuous Variables",
     col = "steelblue",
     pch = 20)






```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
library(ggplot2)

# Scatter plot: WHOLE vs VOLUME, color by CLASS
ggplot(mydata, aes(x = VOLUME, y = WHOLE, color = CLASS)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "WHOLE Weight vs. VOLUME by CLASS",
    x = "VOLUME",
    y = "WHOLE Weight",
    color = "CLASS"
  ) +
  theme_minimal()


```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
max_ratio <- max(mydata$SHUCK / mydata$WHOLE, na.rm = TRUE)
library(ggplot2)

ggplot(mydata, aes(x = WHOLE, y = SHUCK, color = CLASS)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = max_ratio, intercept = 0, color = "black", linetype = "dashed", size = 1) +
  labs(
    title = "SHUCK vs WHOLE with Max SHUCK/WHOLE Ratio Line",
    x = "WHOLE Weight",
    y = "SHUCK Weight",
    color = "CLASS"
  ) +
  theme_minimal()


```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: In the WHOLE vs. VOLUME plot, the points show a very tight linear relationship - as shell volume increases, whole weight increases proportionally across all classes. The spread of points around the trend is relatively small, indicating low variability for a given volume.

In the SHUCK vs. WHOLE plot, the relationship is still linear but shows greater variability for larger whole weights. This makes sense because SHUCK weight is only a part of WHOLE weight, influenced by meat yield, which varies between individuals. For the same whole weight, the shuck weight can differ more depending on the proportion of edible meat.

Regarding age classes (A1–A5): In WHOLE vs. VOLUME, classes are spread along the growth line with older classes (A4, A5) concentrated at higher weights/volumes. In SHUCK vs. WHOLE, the separation is less distinct; variability in meat proportion blurs the distinctions between age classes, especially in the upper weight range.

So overall, the shuck plot shows more vertical scatter and less crisp separation between age classes compared to the whole weight vs. volume plot, reflecting biological variability in meat yield.***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
# Set up 3x3 plotting area
par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))

# Get list of unique SEX values
sex_levels <- levels(mydata$SEX)

# Row 1: Histograms by SEX
for (s in sex_levels) {
  hist(mydata$RATIO[mydata$SEX == s],
       main = paste("Histogram - SEX:", s),
       xlab = "RATIO", col = "skyblue", border = "white")
}

# Row 2: Boxplots by SEX
for (s in sex_levels) {
  boxplot(mydata$RATIO[mydata$SEX == s],
          main = paste("Boxplot - SEX:", s),
          horizontal = TRUE, col = "lightgreen")
}

# Row 3: Q-Q Plots by SEX
for (s in sex_levels) {
  qqnorm(mydata$RATIO[mydata$SEX == s],
         main = paste("Q-Q Plot - SEX:", s))
  qqline(mydata$RATIO[mydata$SEX == s], col = "red")
}


```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: All three groups show slightly right-skewed distributions. The bulk of the data is centered near ~0.13–0.15, with a longer tail extending toward higher ratios. A perfectly normal distribution would be symmetric and bell-shaped in the histogram, and the boxplot’s median line would be near the center with roughly equal whisker lengths. Here, the histograms deviate from perfect bell-shapes, and the boxplots show longer right whiskers and several outliers on the high end. The spread is fairly similar among the sexes, though the I group seems to have slightly less skew than F or M. All groups have upper-end outliers. These suggest occasional individuals with unusually high meat-to-volume ratios. None of the groups perfectly follows a normal distribution. The skewness and presence of outliers make them moderately non-normal.***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
library(dplyr)
library(knitr)

# Initialize empty list to collect outliers for each SEX level
outlier_rows <- list()

# Loop over each SEX category
for (s in levels(mydata$SEX)) {
  
  # Subset the data for this SEX
  subset_data <- mydata %>% filter(SEX == s)
  
  # Get RATIO outliers using boxplot.stats()
  outlier_vals <- boxplot.stats(subset_data$RATIO)$out
  
  # Find the rows in original data that match these outliers
  outlier_rows[[s]] <- subset_data %>% filter(RATIO %in% outlier_vals)
}

# Combine all outlier rows into one data frame
all_outliers <- bind_rows(outlier_rows)

# Display using kable
kable(all_outliers, caption = "Outlying RATIO Values by SEX")


```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer:
1. High RATIO means a high proportion of meat to shell volume
Most of these outliers have RATIO values > 0.21, compared to the overall central tendency (\~0.14). These individuals are unusually “meaty” relative to their shell volume, which could indicate biological differences (age, sex-specific traits, or growth anomalies).
2. Spread across all sexes, but patterns differ
Female: Several outliers in A3 (RINGS = 9–10) and one extreme high value at 0.3118 in class A2, significantly above typical ratios.
Infant: Many extreme values in A1 and A2, with very small shell volumes (some < 20), yet high meat proportion — possibly due to measurement effects on small individuals (ratio inflation from small denominators). 
Male: Mostly large-bodied individuals (VOLUME > 240) with high ratios, suggesting well-developed soft tissue relative to shell size.
3. Influence of shell size on RATIO outliers
Several outliers (especially in I) come from very small shell volumes, where even small changes in SHUCK weight lead to large ratio changes. Larger volume outliers (mainly M and some F) seem genuinely meat-heavy, not just mathematically inflated.
4. Age class distribution
Outliers occur in younger classes (A1, A2) for I and F, but in A3 for M and F — possibly indicating that as some individuals age, certain sexes maintain higher meat proportions. No extreme RATIO outliers from A4 or A5, suggesting older abalones may have lower meat-to-shell proportions.***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
library(ggplot2)
library(gridExtra)

# Plot 1: Boxplot of VOLUME by CLASS
p1 <- ggplot(mydata, aes(x = CLASS, y = VOLUME, fill = CLASS)) +
  geom_boxplot() +
  labs(title = "VOLUME by CLASS", y = "VOLUME", x = "") +
  theme_minimal() +
  theme(legend.position = "none")

# Plot 2: Boxplot of WHOLE by CLASS
p2 <- ggplot(mydata, aes(x = CLASS, y = WHOLE, fill = CLASS)) +
  geom_boxplot() +
  labs(title = "WHOLE by CLASS", y = "WHOLE", x = "") +
  theme_minimal() +
  theme(legend.position = "none")

# Plot 3: Scatterplot of VOLUME vs RINGS
p3 <- ggplot(mydata, aes(x = RINGS, y = VOLUME)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  labs(title = "VOLUME vs RINGS", x = "RINGS", y = "VOLUME") +
  theme_minimal()

# Plot 4: Scatterplot of WHOLE vs RINGS
p4 <- ggplot(mydata, aes(x = RINGS, y = WHOLE)) +
  geom_point(color = "darkgreen", alpha = 0.6) +
  labs(title = "WHOLE vs RINGS", x = "RINGS", y = "WHOLE") +
  theme_minimal()

# Arrange all plots in a 2x2 grid
grid.arrange(p1, p2, p3, p4, nrow = 2)


```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: 
1. General trends
Both VOLUME and WHOLE weight increase with RINGS up to around 10–12 rings, after which the relationship flattens or becomes more scattered. The early growth phase is clear: young abalones (low RINGS) tend to have small volumes and low whole weights. After maturity, the growth in both variables slows, and variability increases significantly.
2. Predictive power by age range
For younger abalones (RINGS ≤ 10–12): Strong relationship - both VOLUME and WHOLE would likely predict age well. For older abalones (RINGS > 12): Weakening relationship - larger variation means predictions would be less accurate. At the same RINGS value, VOLUME and WHOLE can differ widely.
3. Differences between VOLUME and WHOLE
Patterns are very similar, because WHOLE is strongly correlated with VOLUME (both depend on size). VOLUME might capture age-related shell growth slightly better than WHOLE, because WHOLE includes meat weight, which can fluctuate more due to health, season, or reproductive cycle.
Overall assessment
Good predictors for young/mid-aged abalones. Less reliable for older abalones, due to biological variation and plateau in growth. Likely better performance in a nonlinear model (e.g., polynomial or piecewise regression) rather than a straight linear model.***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
library(knitr)

# Step 1: Compute means using aggregate()
means_df <- aggregate(cbind(VOLUME, SHUCK, RATIO) ~ SEX + CLASS, data = mydata, FUN = mean)

# Step 2: Create matrix for VOLUME
volume_mat <- matrix(
  means_df$VOLUME,
  nrow = length(unique(mydata$SEX)),
  ncol = length(unique(mydata$CLASS)),
  byrow = FALSE,
  dimnames = list(
    levels(mydata$SEX),
    levels(mydata$CLASS)
  )
)

# Step 3: Create matrix for SHUCK
shuck_mat <- matrix(
  means_df$SHUCK,
  nrow = length(unique(mydata$SEX)),
  ncol = length(unique(mydata$CLASS)),
  byrow = FALSE,
  dimnames = list(
    levels(mydata$SEX),
    levels(mydata$CLASS)
  )
)

# Step 4: Create matrix for RATIO
ratio_mat <- matrix(
  means_df$RATIO,
  nrow = length(unique(mydata$SEX)),
  ncol = length(unique(mydata$CLASS)),
  byrow = FALSE,
  dimnames = list(
    levels(mydata$SEX),
    levels(mydata$CLASS)
  )
)

# Step 5: Present using kable()
kable(volume_mat, caption = "Mean VOLUME by SEX and CLASS")
kable(shuck_mat, caption = "Mean SHUCK by SEX and CLASS")
kable(ratio_mat, caption = "Mean RATIO by SEX and CLASS")


```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
library(ggplot2)
library(gridExtra)
library(dplyr)

# Step 1: Compute group means
group_means <- mydata %>%
  group_by(CLASS, SEX) %>%
  summarise(
    mean_ratio = mean(RATIO, na.rm = TRUE),
    mean_volume = mean(VOLUME, na.rm = TRUE),
    mean_shuck = mean(SHUCK, na.rm = TRUE),
    .groups = "drop"
  )

# Step 2: Plot mean RATIO vs CLASS
p1 <- ggplot(group_means, aes(x = CLASS, y = mean_ratio, group = SEX, color = SEX)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(title = "Mean RATIO vs CLASS", y = "Mean RATIO") +
  theme_minimal()

# Step 3: Plot mean VOLUME vs CLASS
p2 <- ggplot(group_means, aes(x = CLASS, y = mean_volume, group = SEX, color = SEX)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(title = "Mean VOLUME vs CLASS", y = "Mean VOLUME") +
  theme_minimal()

# Step 4: Plot mean SHUCK vs CLASS
p3 <- ggplot(group_means, aes(x = CLASS, y = mean_shuck, group = SEX, color = SEX)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(title = "Mean SHUCK vs CLASS", y = "Mean SHUCK") +
  theme_minimal()

# Step 5: Arrange all 3 plots
grid.arrange(p1, p2, p3, nrow = 3)


```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: 
The Mean RATIO (SHUCK / VOLUME) decreases with age (A1 → A5) for all sexes.
Does the decreasing ratio indicate that shell growth outpaces meat growth as abalones age? Why do females and males have consistently higher ratios than infant abalones in early classes but converge in later classes?

F and M abalones have higher mean volumes than I abalones in every age class, with females slightly ahead of males at most stages.
Is the volume difference between F and M statistically significant within each class? Does the consistently lower volume of infant abalones mean they mature later, or does it reflect different growth rates?

Mean SHUCK weight rises with class but plateaus or slightly declines at A4–A5, especially in immature abalones.
Does the plateau reflect biological limits to meat growth after a certain age? Is the sharper decline for infant abalones due to stunted development or sample size effects?

Across all plots, growth slows with age for both VOLUME and SHUCK, suggesting diminishing returns in using size to predict age in older abalones.
Would combining VOLUME and RATIO provide a better age predictor than either alone? Are there sex-specific aging curves that should be modeled separately for improved prediction?***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
library(ggplot2)
library(gridExtra)
library(dplyr)

# Filter for RINGS < 16
data_sub <- subset(mydata, RINGS < 16)

# Separate by sex group
infants <- filter(data_sub, SEX == "I")
adults <- filter(data_sub, SEX %in% c("F", "M"))

# Plot 1: VOLUME by RINGS for Infants
p1 <- ggplot(infants, aes(x = factor(RINGS), y = VOLUME)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Infants: VOLUME by RINGS", x = "RINGS", y = "VOLUME") +
  ylim(0, 1100) +
  theme_minimal()

# Plot 2: VOLUME by RINGS for Adults
p2 <- ggplot(adults, aes(x = factor(RINGS), y = VOLUME)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Adults: VOLUME by RINGS", x = "RINGS", y = "VOLUME") +
  ylim(0, 1100) +
  theme_minimal()

# Plot 3: WHOLE by RINGS for Infants
p3 <- ggplot(infants, aes(x = factor(RINGS), y = WHOLE)) +
  geom_boxplot(fill = "pink") +
  labs(title = "Infants: WHOLE by RINGS", x = "RINGS", y = "WHOLE") +
  ylim(0, 400) +
  theme_minimal()

# Plot 4: WHOLE by RINGS for Adults
p4 <- ggplot(adults, aes(x = factor(RINGS), y = WHOLE)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Adults: WHOLE by RINGS", x = "RINGS", y = "WHOLE") +
  ylim(0, 400) +
  theme_minimal()

# Arrange in a 2x2 grid
grid.arrange(p1, p2, p3, p4, nrow = 2)


```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: 
1. Growth patterns over rings (age)
Volume and Whole weight both increase with RINGS up to a certain point and then level off. For adults, the growth curve rises more steeply and plateaus at higher values than for infants. This suggests that once abalones mature, they continue to gain size for several years before growth slows substantially.

2. Infants vs. Adults: Volume
Infants reach a plateau in volume around 10–12 rings, with median volumes significantly lower than adults at the same ring counts. Adults continue to show volume increases into the low teens in ring counts and plateau at much higher medians. 

3. Infants vs. Adults: Whole weight
Infants hit a weight plateau much earlier (around 9–10 rings) and at much lower weight levels. Adults achieve greater whole weights for the same number of rings, and the spread (variability) is much larger, possibly reflecting more diverse environmental/reproductive histories.

4. Differences in variability
Adults show greater variation in both volume and whole weight within the same ring counts. This could mean adult abalones are subject to more diverse living conditions, food availability, or reproductive status. Infants are more tightly clustered in size for a given age, which could indicate more uniform conditions in early growth stages.***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: 
1. Weak correlation between predictors and age
The plots and summaries suggest that although some physical measurements (like length, diameter, height, and weight) generally increase with age, there is considerable overlap between age classes. This overlap reduces the predictive power of these variables.

2. High within-class variability
Younger abalones and older abalones often show similar sizes and weights, likely due to biological variation, sex differences, and environmental factors. This intra-class variation makes classification more difficult.

3. Nonlinear growth patterns
Growth in abalones is not constant over time. Physical growth tends to plateau in older abalones, meaning that beyond a certain age, additional years add little change to size or weight. This creates a many-to-one mapping from measurement to age.

4. Sex differences complicating prediction
Males, females, and infants may follow different growth trajectories, and sex-specific differences can mask the relationship between size and age unless explicitly modeled.

5. Potential measurement limitations
Variables like “shuck weight” and “ratio” show high variability even within the same age groups. Since SHUCK is part of WHOLE, these variables are correlated and may not provide much independent predictive information.

6. Class imbalance
The data distribution by age class (RINGS/CLASS) is highly uneven, with some ages being rare. This imbalance can bias models toward predicting more common age groups, reducing accuracy for less common ones.

In short, the physical measurements available capture some growth trends but lack the precision to uniquely determine age, especially for mid-to-late life stages where growth slows. This inherent statistical limitation likely explains the poor predictive performance.***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: 
1. About the data collection process
* How was the sample collected? Was it random, stratified, or convenience sampling?
* What is the sample size, and is it adequate for capturing variability?
* Was the sampling method consistent across all observations?
* Over what time period and in what context were the data collected? Could seasonality or temporal changes affect results?

2. About data quality and measurement
* How were measurements taken? Are the instruments or procedures reliable?
* Were there any known sources of measurement error or bias?
* Are there missing values, and if so, how were they handled?

3. About representativeness and bias
* Does the sample cover all relevant subgroups of the population?
* Could certain subgroups be over- or under-represented?
* Were there any inclusion or exclusion criteria that might skew results?

4. About the summary statistics themselves
* Are there extreme outliers that might distort the mean or variance?
* Do measures of central tendency (mean, median) tell the same story, or are they far apart?
* Is the distribution skewed or symmetric? Does the histogram suggest multiple modes (possible subpopulations)?

5. About the context and interpretation
* Is there prior knowledge or a theoretical distribution to compare with?
* Could the observed pattern be due to random variation alone?
* Have any transformations or aggregations been applied that might obscure details?***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer:
1. Lack of control over variables
In observational studies, the researcher does not control who is exposed to what conditions or treatments. Many variables may influence the outcome simultaneously, making it hard to isolate the effect of any single factor.

2. Confounding variables
Unmeasured or uncontrolled factors may influence both the independent and dependent variables. Confounding can lead to false associations (spurious relationships) or mask real ones.

3. Selection bias
Participants are not randomly assigned. Certain groups may be over- or under-represented, affecting representativeness and validity.

4. Measurement bias
Data may be inconsistently or inaccurately measured, especially if collected by different observers, at different times, or under different conditions.

5. Temporal ambiguity
The sequence of events may be unclear — did the suspected cause occur before the effect? Without a clear timeline, determining directionality is challenging.

Causality in observational studies
Direct causality generally cannot be proven from observational studies alone because of uncontrolled confounding and the absence of experimental manipulation. Statistical methods can help control for confounding, but causal inference still requires strong assumptions and supporting evidence. At best, observational studies can suggest possible causal relationships, which can then be tested more rigorously in controlled experiments.

What can be learned from observational studies
Association patterns: Identifying correlations between variables that may warrant further study.
Natural variation: Understanding real-world conditions, diversity, and variability in the population.
Hypothesis generation: Providing the basis for designing future experiments.
Long-term or unethical-to-manipulate questions: For example, effects of smoking on health, where experimental exposure would be unethical.
Prevalence and trends: Estimating how common certain conditions, behaviors, or outcomes are in different populations.***
